Categorical Foundations of E-Graphs With Binding

### 2025-07-09 POPL2026/paper/sections/lambda-calculus.tex
mostly finish lambda calculus section

\emph{Graphical equivalence} $\sim$ of $\lambda_{\text{lsub}}$-terms is augmentation of $\alpha$-equivalence of terms with the additional stipulations that
\begin{align*}
	t [x/u] [y/v] \quad        & \sim \quad t [y/v] [x/u],        & x \notin \textsf{fv} (v) \land y \notin \textsf{fv} (u), \\
	(\lambda y. t) [x/u] \quad & \sim \quad \lambda y. (t [x/u]), & y \notin \textsf{fv} (u),                                \\
	(t v) [x/u] \quad          & \sim \quad t [x/u] v,            & x \notin \textsf{fv} (u),
\end{align*}
closed under contexts, transitivity, symmetry, and reflexivity.
Put simply, it is the stipulation that \textbf{let}-bindings may be freely reordered without changing the meaning of the term\footnote{Recall that our setting is \emph{pure}, in the sense that no sequencing of effects is meaningful.}.
The $\lambda^\sim_{\text{lsub}}$-calculus is given by these graphical equivalence classes of terms with the reduction relation defined by that of the $\lambda_{\text{lsub}}$-calculus modulo graphical equivalence.

This calculus is explicitly designed to capture the representation of terms as proof nets: for two terms $t, u$ in the $\lambda_{\text{lsub}}$-calculus, $t \sim u$ if and only if both map to the same proof net~\cite{?}.
It turns out that the same fact is true for the representation of terms as string diagrams~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} instead of proof terms, and in this fact we can exploit the string diagram setting of e-hypergraphs to absorb the reductions arising due to graphical equivalence into our representation rather than explicitly providing them as equations attached to a symmetric monoidal theory presenting the $\lambda^\sim_{\text{lsub}}$-calculus.

@@

We will define a monoidal signature whose closed symmetric monoidal theory is the $\lambda^\sim_{\text{lsub}}$-calculus extended by tupling of terms and flat natural numbers equipped with basic arithmetic operations.
Let $\Sigma \coloneq (\{ \mathbf{N} \}, \{ \comonoid, \counit \} \cup \mathbb{N} \cup \{ \text{plus}, \text{mult} \}, t)$ where
\begin{align*}
	t\colon \{ \comonoid, \counit \} \cup \mathbb{N} \cup \{ \text{plus}, \text{mult} \} & \to {\{ \mathbf{N} \}}^* \times {\{ \mathbf{N} \}}^*                                                    \\
	t(n)                                                                                 & = ((), (\mathbf{N})),                                & n \in \mathbb{N},                                \\
	t(o)                                                                                 & = ((\mathbf{N}, \mathbf{N}), (\mathbf{N})),          & o \in \{ \comonoid, \text{plus}, \text{mult} \}, \\
	t(\counit)                                                                           & = ((\mathbf{N}), ()).
\end{align*}
The operations that the signature $\Sigma$ avails are copy and delete, denoted by $\comonoid$ and $\counit$ respectively, constants for each natural number, and the binary arithmetic operations $+$ and $\times$.
We interpret the $\Sigma$-terms $\llbracket \tau \rrbracket$ as follows: fully applied operations as themselves, $\id_{I}$ and $\id_{\mathbf{N}}$ as the identity function $\lambda x. x$ on the empty tuple $()$ and $\mathbf{N}$ respectively, $\sym_{\mathbf{N}, \mathbf{N}}$ as the swap function $\lambda (x, y). (y, x)$, composition $\tau_1 ; \tau_2$ as the composition function $\lambda x. \llbracket \tau_2 \rrbracket \left(\llbracket \tau_1 \rrbracket x\right)$, and $\tau_1 \otimes \tau_2$ as tupling $\left(\llbracket \tau_1 \rrbracket, \llbracket \tau_2 \rrbracket\right)$.

@@

>>>>

\emph{Graphical equivalence} $\sim$ of $\lambda_{\text{lsub}}$-terms is augmentation of $\alpha$-equivalence of terms with the additional stipulations that
\begin{align*}
	t [x/u] [y/v] \quad        & \sim \quad t [y/v] [x/u],        & x \notin \textsf{fv} (v) \land y \notin \textsf{fv} (u), \\
	(\lambda y. t) [x/u] \quad & \sim \quad \lambda y. (t [x/u]), & y \notin \textsf{fv} (u),                                \\
	(t v) [x/u] \quad          & \sim \quad t [x/u] v,            & x \notin \textsf{fv} (v),
\end{align*}
closed under contexts, transitivity, symmetry, and reflexivity.
Put simply, it is the stipulation that \textbf{let}-bindings may be freely reordered without changing the meaning of the term\footnote{Recall that our setting is \emph{pure}, in the sense that no sequencing of effects is meaningful.}.
The $\lambda^\sim_{\text{lsub}}$-calculus is given by these graphical equivalence classes of terms with the reduction relation defined by that of the $\lambda_{\text{lsub}}$-calculus modulo graphical equivalence.

This calculus is explicitly designed to capture the representation of terms as proof nets: for two terms $t, u$ in the $\lambda_{\text{lsub}}$-calculus, $t \sim u$ if and only if both map to the same proof net~\cite{?}.
It turns out that the same fact is true for the representation of terms as string diagrams~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} instead of proof nets, and in this fact we can exploit the string diagram setting of e-hypergraphs to absorb the reductions arising due to graphical equivalence into our representation rather than explicitly providing them as equations attached to a closed symmetric monoidal theory presenting the $\lambda^\sim_{\text{lsub}}$-calculus.

@@

We will define a monoidal signature whose closed symmetric monoidal theory is the $\lambda^\sim_{\text{lsub}}$-calculus extended by tupling of terms and flat natural numbers equipped with basic arithmetic operations.
Let $\Sigma \coloneq (\{ \mathbf{N} \}, \{ \comonoid, \counit \} \cup \mathbb{N} \cup \{ \text{plus}, \text{mult} \}, t)$ where
\begin{align*}
	t\colon \{ \comonoid, \counit \} \cup \mathbb{N} \cup \{ \text{plus}, \text{mult} \} & \to {\{ \mathbf{N} \}}^* \times {\{ \mathbf{N} \}}^*                                                    \\
	t(n)                                                                                 & = ((), (\mathbf{N})),                                & n \in \mathbb{N},                                \\
	t(o)                                                                                 & = ((\mathbf{N}, \mathbf{N}), (\mathbf{N})),          & o \in \{ \comonoid, \text{plus}, \text{mult} \}, \\
	t(\counit)                                                                           & = ((\mathbf{N}), ()).
\end{align*}
The operations that the signature $\Sigma$ avails are copy and delete, denoted by $\comonoid$ and $\counit$ respectively, constants for each natural number, and the binary arithmetic operations $+$ and $\times$.
We interpret the $\Sigma$-terms $\llbracket \tau \rrbracket$ as follows: operations as themselves, $\id_{I}$ and $\id_{\mathbf{N}}$ as the identity function $\lambda x. x$ on the empty tuple $()$ and $\mathbf{N}$ respectively, $\sym_{\mathbf{N}, \mathbf{N}}$ as the swap function $\lambda (x, y). (y, x)$, composition $\tau_1 ; \tau_2$ as the composition function $\lambda x. \llbracket \tau_2 \rrbracket \left(\llbracket \tau_1 \rrbracket x\right)$, and $\tau_1 \otimes \tau_2$ as tupling $\left(\llbracket \tau_1 \rrbracket, \llbracket \tau_2 \rrbracket\right)$.

@@

\begin{figure}
	\centering
	\begin{subfigure}{0.3\linewidth}
		\begin{lstlisting}[language=ML]
let x = 1 in
let y = 2 in
plus(x, y)
                \end{lstlisting}
		\hrule
		\begin{lstlisting}[language=ML]
let y = 2 in
let x = 1 in
plus(x, y)
                \end{lstlisting}
		\hrule
		\begin{lstlisting}[language=ML]
plus(1, 2)
                \end{lstlisting}
		\caption{Three programs for $1 + 2$}%
		\label{fig:plus-1-2-programs}
	\end{subfigure}
	\hspace{2em}
	\begin{subfigure}{0.2\linewidth}
		\begin{align*}
			     & \text{plus}(x, y)[y/2][x/1] \\
			\sim &                             \\
			     & \text{plus}(x, y)[x/1][y/2] \\
			\sim &                             \\
			     & \text{plus}(1, 2)
		\end{align*}
		\caption{As terms in $\lambda_{\text{lsub}}$}%
		\label{fig:plus-1-2-terms}
	\end{subfigure}
	\hspace{2em}
	\begin{subfigure}{0.2\linewidth}
		\centering
		\tikzfig{figures/plus-1-2}
		\caption{String diagrammatic representation}%
		\label{fig:plus-1-2-string-diagram}
	\end{subfigure}
	\caption{Graphically equivalent terms}%
	\label{fig:plus-1-2}
\end{figure}

\autoref{fig:plus-1-2-programs} shows some programs which are graphically and semantically equivalent but syntactically distinct.
As alluded to in~\autoref{sec:vs-e-graphs-with-binding}, the traditional (slotted) e-graph approach to modelling the graphical equivalence of these terms (\autoref{fig:plus-1-2-terms}) relies on explicitly encoding substitution operations as e-graph nodes, and their behaviour (the graphical equivalence relation $\sim$) as rewrite rules, as in~\autoref{fig:e-graph-substitution} --- not only is this more complicated, but it also leads to a blow-up in the state space required to perform equality saturation in any extension of the $\lambda^\sim_{\text{lsub}}$-calculus, because we are really rewriting $\lambda_{\text{lsub}}$-terms and computational resources are spent rewriting between graphically equivalent terms.
\textcolor{red}{Aleksei, can you say something quantitative here if you ran some tests?}
In contrast, in our framework, $\lambda^\sim_{\text{lsub}}$-terms are directly modelled as string diagrams (\autoref{fig:plus-1-2-string-diagram}\footnote{Note that dashed boxes are elided for convenience here because no rewriting has occurred.}).
We do however note that the traditional approach is more flexible, in the sense that it may model more theories with binders (e.g.\ $\pi$-calculus), whereas our approach is essentially specialised to $\lambda$-calculus, relying heavily on the metatheoretic internal language of the ambient setting.

As a second example, we may wish to incorporate the optimisation ${\text{plus} (x, x) \leadsto \text{mult} (x, 2)}$ in our toy theory.
This corresponds to endowing $\mathcal{E}$ with the equation of string diagrams:
\[
	\tikzfig{figures/plus-copy}
	\quad
	=
	\quad
	\tikzfig{figures/mult-2}
	\;
	.
\]
Now, in our semantic model, this rewrite can be non-destructively applied to model one step of e-graph saturation:
\[
	\tikzfig{figures/plus-copy-rewrite-1}
	\quad
	\equiv
	\quad
	\tikzfig{figures/plus-copy-rewrite-2}
	\quad
	\to
	\quad
	\tikzfig{figures/plus-copy-rewrite-3}
	\;
	.
\]
Here, $\equiv$ is a definitional equality arising from the idempotence in the Hom semilattice, and $\to$ is an application of DPO rewriting using the previous equation along the right branch.
In isolation, this isn't particularly interesting, but the important observation is that this rewriting all happens \emph{locally} in the context of a bigger diagram.

From a practical perspective, the encoding of string diagrams as (e-)hypergraphs provides a normal form for graphical equivalence classes of terms.
As a data structure, this provides an efficient representation upon which to build an implementation of this theory as a fully fledged \enquote{e-graphs for string diagram rewriting}, using DPO rewriting as a primitive to implement the standard e-graph operations.
We leave this as future work, noting its applications to not only $\lambda^\sim_{\text{lsub}}$-calculus, but also other monoidal theories where there are two levels of rewriting:
\begin{enumerate}
	\item \enquote{bureaucratic} reassociation and reparenthesisation, absorbed into isotopy of string diagrams (here, graphical equivalence of $\lambda_{\text{lsub}}$-terms);
	\item algebraic identities arising from the semantics of the theory (e.g. ${\text{plus} (x, x) \leadsto \text{mult} (x, 2)}$ for arithmetic).
\end{enumerate}
One such theory is the ZX calculus~\cite{coeckeInteractingQuantumObservables2011}.
